Strength, mating success, and immune and energy costs in a population sample of US women and men: a registered report

### 2024-08-23 paper.qmd
Latest edits

---
title: "Grip strength, mating success, and immune and energy costs in a population sample of US women and men: a registered report"
author: 
  - name: Caroline B. Smith
    affiliation:
     - id: wsu
       department: Department of Anthropology
       name: Washington State University
  - name: Edward H. Hagen
    affiliation:
     - ref: wsu
date: today
lightbox: auto
format:
  html:
    toc: true
    embed-resources: true
  # docx:
  #   number-sections: true
  #   fig-format: pdf
execute:
  echo: false
  warning: false
csl: apa.csl
bibliography: paper.bib
crossref:
  custom:
    - kind: float
      reference-prefix: Figure S
      key: sfig
      space-before-numbering: false
    - kind: float
      reference-prefix: Table S
      key: stbl
      space-before-numbering: false
---

@@

According to the sexual selection hypothesis, males evolved greater physical formidability than females because physical formidability increased the mating success of males more than females. The aim of this study is to replicate the costs and benefits of formidability reported by @lassek2009 using similar nationally representative data, controlling for a wider range of possible confounds, and precisely prespecifying the statistical models in a registered report format. It also aims to investigate if these costs and benefits are also experienced by women, and specifically if the effect of physical strength on mating success, a proxy for reproductive success, is greater for men than women, as the sexual selection hypothesis predicts.

@@

Our methods and models remained identical to those reported in the pilot study, other than the deviation reported below, using unobserved data from the 2013-2014 NHANES collection cycle. After testing and reporting our predictions using the unobserved data, we fit our models on the combined data (2011-2014), and conduct other post hoc analyses.

## Deviation from Stage 1 method

Models and R code for three of our four mating success outcome variables remain identical to those used in the Stage 1 pilot study. However, although the Poisson regression model of lifetime sexual partners in the pilot study included age as a control, it is more appropriate to instead include age as an exposure variable (an offset) in the model, which we now do. We also treat this exposure as years since sexual maturity (age - 12 years). The changes in the strength and $strength\times sex$ coefficients from this change in model specification are minimal (see @sfig-comparison).

As described in the methods section, a different set of weights is required in models that utilize dietary data. In one model of pilot data (the expanded controls immune model) we failed to do so, an oversight we correct here in both the pilot and confirmatory models.

@@

# plot_coefs + plot_coefs_stage2 + plot_layout(axes = 'collect_y', axis_titles = 'collect')
plot_coefs_combined
```
Our confirmatory results revealed strong positive significant main effects of sex-specific strength on partnered status, numbers of lifetime partners after controlling for all five sets of potential confounds. We found mixed evidence for a main effect of strength in the age of first sex models, as strength was not a significant predictor after controlling for socioeconomic, anthropometric, or physical activity controls. Strength was not a predictor of past year number of sexual partners in any control model.

@@

```{r}
#| label: fig-immune
#| fig.width: 12
#| fig.height: 6
#| fig.cap: Coefficient plot of predictors of immune investment (WBCC) from generalized linear models (Pilot). Variables labeled (S) have been centered at the mean and standardized by 2 SD. Bars are 95% CIs.

# plot_immune_both <- 
#   plot_immune + guides(colour = 'none', shape = 'none') + plot_immune_stage2 + plot_layout(axes = 'collect', guides = 'collect')

@@

# p <- plot_energy_coefs / plot_protein_coefs + plot_layout(guides = 'collect', axis_titles = 'collect')
ggsave("Figures/plot_intake_strength_combined.pdf", plot_intake_strength_combined, width = 12, height = 6)
plot_intake_strength_combined
```

## Discussion
The purpose of this study was to replicate as closely as possible, the results from @lassek2009, in new data that used combined grip strength as a proxy for upper body strength. Unlike @lassek2009, we included women to test if the effects of upper body strength on mating success were greater for men than women, as predicted by the sexual selection hypothesis. We also included numerous additional control variables, and used a registered report format with pre-specified models developed using pilot data in Stage 1 and then fit on held-out data in Stage 2.

In the Stage 2 Confirmatory Study, we found consistent support for a positive relationship between combined grip strength and two measures of mating success: being currently partnered, and lifetime number of sexual partners for men, controlling for numerous possible confounds. We found no consistent association between strength and age at first sex or number of past year sexual partners.

@@

The lack of a sex difference in the significant positive effect of sex-specific strength on short term mating success (proxied by number of lifetime sexual partners) is puzzling. Because men have greater strength than women their self-reported partner numbers are higher, on average. Still, women with higher grip strength report more lifetime sexual partners than women with lower grip strength even after controlling for a host of potential confounds.

It could be that there was selection for more formidable men to prefer more partner variety, and stronger women have a similar preference as a byproduct of selection on men. It could be that there is assortative mating on strength; thus if stronger men are motivated to switch partners more frequently, their (stronger) mates would also likely have more mates. It might be that stronger women require less male investment and so instead benefit from greater partner numbers through, e.g. genetic bet-hedging, forging relationships with multiple males, ability to conduct a more extensive search for a high quality long term mate, or through avoiding costly long term partnerships. It might also be the case that there are some sex-specific confounds that we failed to control for. Greater partner numbers might indicate mating failures rather than mating successes for males or females, although why strength would be associated with mating failures is not clear. Since this sex difference was significant in the Pilot Study (2011-12), but not in the Confirmatory Study (2013-14), it is possible that cultural changes influencing sexual behavior occurred between the data collection cycles.

@@

We thank the Editor and two anonymous reviewers for their numerous helpful comments and suggestions.

@@

:::{#stbl-power}
```{r}
v <- ifelse(as.matrix(df_pwr[-1])>0.8, T, F)
tt(df_pwr) |>
  group_tt(j = list("Partnered" = 2:3, "Lifetime partners" = 4:5, "Last year partners" = 6:7)) |>
  format_tt(j = 1:7, digits = 2, num_zero = T) |>
  style_tt(j = 1:7, align = c("ccccccc")) |> 
  style_tt(i=1:4, j=2:7, bold = v)
```

@@

<!-- ### Models of mating success with anthropometric, socioeconomic, health, hormone, and physical activity controls -->

@@

:::{#sfig-sexspecific}
```{r}
#| fig-width: 9
#| fig-height: 9
plot_sexspecific_strength_coefs
```

@@

:::{#sfig-comparison}
```{r}
#| fig-width: 10
plot_compare_age_models
```

@@

:::{#sfig-sexspecific-confirm}
```{r}
#| fig-width: 9
#| fig-height: 9
plot_sexspecific_strength_coefs_confirm
```

@@

:::{#sfig-matingGH}

@@

:::{#sfig-effectsGH}

@@

:::{#sfig-immunityGH}

@@

:::{#sfig-intakeGH}

@@

>>>>

---
title: "Strength, mating success, and immune and energy costs in a population sample of US women and men: a registered report"
author: 
  - name: Caroline B. Smith
    affiliation:
     - id: wsu
       department: Department of Anthropology
       name: Washington State University
  - name: Edward H. Hagen
    affiliation:
     - ref: wsu
date: today
lightbox: auto
format:
  html:
    toc: true
    embed-resources: true
  # docx:
  #   number-sections: true
  #   fig-format: pdf
execute:
  echo: false
  warning: false
csl: apa.csl
bibliography: paper.bib
crossref:
  custom:
    - kind: float
      reference-prefix: Figure S
      key: sfig
      space-before-numbering: false
    - kind: float
      reference-prefix: Table S
      key: stbl
      space-before-numbering: false
---

@@

According to the sexual selection hypothesis, males evolved greater physical formidability than females because physical formidability increased the mating success of males more than females. The aim of this study is to replicate the costs and benefits of formidability reported by @lassek2009 using similar nationally representative data, controlling for a wider range of possible confounds, and precisely prespecifying the statistical models in a registered report format. It also aims to investigate if these costs and benefits are also experienced by women, and specifically if there is a positive effect of physical strength on male mating success, a proxy for reproductive success, that is significantly greater than its effect on women's mating success, as the sexual selection hypothesis predicts. See @fig-sexualselection.

::: {#fig-sexualselection}
![](sexualselection.svg)

Predicted relationships between upper body strength and various indices of mating success, such as numbers of partners, partnered status, and age at first sex, for both men and women. According to the sexual selection hypothesis, sexual dimorphism in upper body strength is explained by a positive effect on male mating success that is greater than its effect on female mating success.
:::

@@

Our methods and models remained identical to those reported in the pilot study, other than the deviations reported below, using unobserved data from the 2013-2014 NHANES collection cycle. After testing and reporting our predictions using the unobserved data, we fit our models on the combined data (2011-2014), and conduct other post hoc analyses.

## Deviations from Stage 1 methods

Models and R code for three of our four mating success outcome variables remain identical to those used in the Stage 1 pilot study. However, although the Poisson regression model of lifetime sexual partners in the pilot study included age as a control, it is more appropriate to instead include age as an exposure variable (an offset) in the model, which we now do. We also treat this exposure as years since sexual maturity (defined this as age - 12 years because 95% of participants first had sex at age 12 or older). The changes  in the strength and $strength\times sex$ coefficients and their standard errors from this change in model specification are minimal (see @sfig-comparison).

As described in the methods section, a different set of weights is required in models that utilize dietary data. In one of the models of pilot data using dietary variables (the expanded controls immune model) we used the MEC weights instead of the dietary weights, an oversight we correct here in both the pilot and confirmatory models.

We downloaded the 2011-2014 NHANES data for this project from the CDC website on Feb 7, 2020. On March 20, 2020, Dual-Energy X-ray Absorptiometry (DXA) data were published for these series, which we unfortunatey failed to notice. These data provide alternatives to grip strength for assessing muscle mass, and include the lean masses (i.e., fat-free mass, optionally including or excluding bone mineral content) in grams of each arm and leg, the trunk, and total body. Total body lean mass is similar to the bioelectrical impedence analysis fat-free mass variable used by @lassek2009. For reviews of DXA measurements of body composition, see @shepherd2017 and @messina2020. For details of the NHANES DXA measurements, see @DXXG2023. In addition to our grip strength models, we also fit our models exactly as specified in stage 1 except substituting four DXA lean masses for grip strength (all masses excluded bone mineral content). Specifically, we computed arm lean mass as the sum of left and right arm lean masses, and leg lean mass as the sum of left and right leg lean masses. We then fit our models using arm, leg, trunk, and total lean masses, each standardized by sex as described earlier for grip strength.

@@

# plot_coefs + plot_coefs_stage2 + plot_layout(axes = 'collect_y', axis_titles = 'collect')
plot_coefs_combined
```

```{r}
#| label: fig-leancoefs
#| fig-width: 12
#| fig-height: 8
#| fig-cap: "Coefficients of sex-specific arm lean mass, and lean mass X sex interaction for our four mating success outcomes and our five sets of potential confounding variables. Partnered coefficients are from logistic regressions, partner number coefficients are from quasi-Poisson glm regressions, and age at first sex coefficients are from Gaussian glm regressions. Bars are 95% CIs."

lean_mass_plot("Arm lean mass", "arm_lean_centered", "arm_lean_sex_centered")
```

Our confirmatory results revealed strong positive significant main effects of sex-specific grip strength and arm lean mass on partnered status and numbers of lifetime partners after controlling for all five sets of potential confounds. We found mixed evidence for a main effect of strength  and lean mass in the age of first sex models, as these were not significant predictors after controlling for socioeconomic, anthropometric, or physical activity controls. Strength and arm lean mass were not predictors of past year number of sexual partners in any control model. Leg, trunk, and total lean mass were not consistently related to any mating success outcome. See @sfig-legs - @sfig-total.

@@

```{r}
#| label: fig-immune
#| fig.width: 12
#| fig.height: 6
#| fig.cap: Coefficient plot of predictors of immune investment (WBCC) from generalized linear models. Variables labeled (S) have been centered at the mean and standardized by 2 SD. Bars are 95% CIs.

@@

ggsave("Figures/plot_intake_strength_combined.pdf", plot_intake_strength_combined, width = 12, height = 6)
plot_intake_strength_combined
```

## Discussion

The purpose of this study was to replicate, as closely as possible, the @lassek2009 study with new data that used combined grip strength as a proxy for upper body strength, as well as the lean masses of arms, legs, trunk, and total body. Unlike @lassek2009, we included women to test if the effects of upper body strength and muscle masses on mating success were greater for men than women, as predicted by the sexual selection hypothesis. We also included numerous additional control variables, and used a registered report format with pre-specified models developed using pilot data in Stage 1 and then fit on held-out data in Stage 2.

In the Stage 2 Confirmatory Study, we found consistent support for a positive relationship between combined grip strength and arm lean mass, and two measures of mating success: being currently partnered, and lifetime number of sexual partners for men, controlling for numerous possible confounds. We found no consistent association between strength or lean masses, and age at first sex or number of past year sexual partners.

@@

The lack of a sex difference in the significant positive effect of sex-specific strength on short term mating success (proxied by number of lifetime sexual partners) is puzzling. Men have greater strength than women and their average partner numbers are higher. Still, women with higher grip strength unexpectedly reported more lifetime sexual partners on average than women with lower grip strength even after controlling for a host of potential confounds. It could be that there was selection for more formidable men to prefer more partner variety, and stronger women have a similar preference as a byproduct of selection on men. It could be that there is assortative mating on strength: if stronger men are motivated to switch partners more frequently, their (stronger) mates would also likely have more mates. It might be that stronger women require less male investment and so instead benefit from greater partner numbers through, e.g. genetic bet-hedging, forging relationships with multiple males, ability to conduct a more extensive search for a high quality long term mate, or through avoiding costly long term partnerships. It might be the case that there are some sex-specific confounds that we failed to control for, or that our models are otherwise misspecified. There also might be reverse causation, e.g., women who are interested in greater partner variety keep in better shape. Greater partner numbers might indicate mating failures rather than mating successes for males or females, although why strength would be associated with mating failures is not clear. Since this sex difference was significant in the Pilot Study (2011-12), but not in the Confirmatory Study (2013-14), it is possible that cultural changes influencing sexual behavior occurred between the data collection cycles. Or these results might simply be noise.

@@

We thank the Editor and two anonymous reviewers for their numerous helpful comments and suggestions. We thank Michael Barlev for pointing us to the DXA data.

@@

::: {#stbl-power}
```{r}
v <- ifelse(as.matrix(df_pwr[-1])>0.8, T, F)
tt(df_pwr) |>
  group_tt(j = list("Partnered" = 2:3, "Lifetime partners" = 4:5, "Last year partners" = 6:7)) |>
  format_tt(j = 1:7, digits = 2, num_zero = T) |>
  style_tt(j = 1:7, align = c("ccccccc")) |> 
  style_tt(i=1:4, j=2:7, bold = v)
```

@@

@@

::: {#sfig-sexspecific}
```{r}
#| fig-width: 9
#| fig-height: 9
plot_sexspecific_strength_coefs
```

@@

::: {#sfig-comparison}
```{r}
#| fig-width: 10
plot_compare_age_models
```

@@

::: {#sfig-sexspecific-confirm}
```{r}
#| fig-width: 9
#| fig-height: 9
plot_sexspecific_strength_coefs_confirm
```

@@

::: {#sfig-matingGH}

@@

::: {#sfig-effectsGH}

@@

::: {#sfig-immunityGH}

@@

::: {#sfig-intakeGH}

@@

## Lean mass coefficient plots

### Arm lean mass

::: {#sfig-arms}

```{r}
#| fig-width: 11
lean_mass_plot("Arm lean mass", "arm_lean_centered", "arm_lean_sex_centered", mating_only = F)
```

Coefficient plots for arm lean mass vs. mating success outcomes, energy and protein intake, and immunity (white blood cell count).
:::

### Leg lean mass

::: {#sfig-legs}

```{r}
#| fig-width: 11
lean_mass_plot("Leg lean mass", "leg_lean_centered", "leg_lean_sex_centered", mating_only = F)
```

Coefficient plots for leg lean mass vs. mating success outcomes, energy and protein intake, and immunity (white blood cell count).
:::

### Trunk lean mass

::: {#sfig-trunk}

```{r}
#| fig-width: 11
lean_mass_plot("Trunk lean mass", "trunk_lean_centered", "trunk_lean_sex_centered", mating_only = F)
```

Coefficient plots for trunk lean mass vs. mating success outcomes, energy and protein intake, and immunity (white blood cell count).
:::

### Total lean mass

::: {#sfig-total}
```{r}
#| fig-width: 11
lean_mass_plot("Total lean mass", "total_lean_centered", "total_lean_sex_centered", mating_only = F)
```

Coefficient plots for total lean mass vs. mating success outcomes, energy and protein intake, and immunity (white blood cell count).
:::
